A \proglang{Python}/\proglang{Zig} optimized and customizable implementation for the $\rho_{DCCA}$ and $DMC_x^2$ methods

### 2025-01-31 paper_v02/article.tex
paper version 02

The \verb"dmcx2_from_p_dcca_matrix_2d()" function uses \pkg{Numpy} methods to preparate the data to apply Eq. \ref{eq:dmcx2}. The list of indexes is divided in \verb"y_index", holding an one item array wiht the index of the dependent variabel time series, and \verb"x_indexes", containing the indexes of the independent ones. The \pkg{Numpy} function \verb"np.ix_()" constructs index arrays that will index the cross product from a serie of 1D arrays as inputs.

>>>>

The \verb"dmcx2_from_p_dcca_matrix_2d()" function uses \pkg{Numpy} methods to preparate the data to apply Eq. \ref{eq:dmcx2}. The list of indexes is divided in \verb"y_index", holding an one item array whit the index of the dependent variabel time series, and \verb"x_indexes", containing the indexes of the independent ones. The \pkg{Numpy} \verb"np.ix_()", althow it's not a very known funcntion of the library constructs index arrays that will use the cross product from a serie of 1D arrays as inputs. It's a very convenient way to extract a submatrix and a vector from the $\rho_{DCCA}$ matrix. The matrix, as assembled by the \verb"p_dcca()" function, will always need to have extracions of a 2D matrix and a vector, as we can see in Eq.~\ref{eq:dmcx2}. The code below is an example of the extraction process.

\begin{CodeChunk}
\begin{CodeInput}
import numpy as np
arr = np.array([[1,2,3,4],
                [2,1,5,6],
                [3,5,1,7],
                [4,6,7,1]])
print(arr)
\end{CodeInput}

\begin{CodeOutput}
[[1 2 3 4]
 [2 1 5 6]
 [3 5 1 7]
 [4 6 7 1]]
\end{CodeOutput}
An illustrative \verb"arr" matrix is defined as $4 \times 4$ with all ones in the main diagonal and symmetric integer values in the other cells. Since the $\rho_{DCCA}$ ranges from $-1$ to $1$, those values should be interpreted as place holders. 
\begin{CodeInput}
sub_mat_index = np.ix_([1,2,3], [1,2,3])
print("index combination:\n", sub_mat_index)
print("extracted matrix:\n",arr[sub_mat_index])
\end{CodeInput}

\begin{CodeOutput}
index combination:
(array([[1],
      [2],
      [3]]), array([[1, 2, 3]]))
extracted matrix:
[[1 5 6]
[5 1 7]
[6 7 1]]
\end{CodeOutput}

Above, a submatrix, holding the positions $1$ to $3$ is extracted using \verb"np.ix_()" function, and below, the extraction of the vector, first as a line and then as a column vector.

\begin{CodeInput}
sub_mat_vec = np.ix_([0], [1,2,3])
print("line vectror:\n", arr[sub_mat_vec])
sub_mat_vec = np.ix_([1,2,3],[0])
print("column vectror:\n", arr[sub_mat_vec])
\end{CodeInput}

\begin{CodeOutput}
line vectror:
[[2 3 4]]
column vectror:
[[2]
[3]
[4]]
\end{CodeOutput}
\end{CodeChunk}

In the \pkg{Zebende} package, the vector is extracted as a column for better coherence with the $DMC_{x}^{2}$ theory. The method also works for dependent variable different of index $0$. The resulting matrix will preserve the diagonal as $1$, the symmetry regarding the main diagonal and the order of elements in the column vector respeted. the code below extract the index $1$ series as the dependent and the others as independent.
\begin{CodeChunk}
\begin{CodeInput}
sub_mat_index = np.ix_([0,2,3], [0,2,3])
print("index combination:\n", sub_mat_index)
print("extracted matrix:\n",arr[sub_mat_index])
sub_mat_vec = np.ix_([0,2,3],[1])
print("column vectror:\n", arr[sub_mat_vec])
\end{CodeInput}

\begin{CodeOutput}
index combination:
(array([[0],
      [2],
      [3]]), array([[0, 2, 3]]))
extracted matrix:
[[1 3 4]
[3 1 7]
[4 7 1]]
column vectror:
[[2]
[5]
[6]]
\end{CodeOutput}
\end{CodeChunk}
